---
title: "Permutation Example"
date: "Sunday, September 07, 2014"
output: html_document
---

```{r knitrsetup, echo=FALSE, message=FALSE, warning=FALSE}
library(knitr)
opts_chunk$set(fig.align = 'center', 
               fig.height = 5, 
               fig.width = 8, 
               tidy = F, 
               cache = F)
```


The packages used and loading the data set.

```{r setup, echo=TRUE, message=FALSE, warning=FALSE}
library(ggplot2)
library(scales)
library(MASS)
library(plot3D)
library(reshape2)
library(mosaic)
library(magrittr)
library(dplyr)
library(tidyr)
set.seed(42)
load("data/LvL_Data.RData")
```

There are two functions in the `permutation_functions.R` script.  `getObsDiff` calculates the observed difference in densities.  You have to pass in the number of grid points in each direction.  It uses the `kde2d` function from the MASS package.

```{r obsdiff, message=FALSE}
source("code/permutation-testing/permutation_functions.R")
n <- 25 # number of grid points in x and y directions
obs_diff2 <- getObsDiff(n)
dim(obs_diff) # one density for each (x,y) grid point

# plot of observed differences
persp3D(x = 1:n, y = 1:n, z = obs_diff)
```

The second function `permuteDensities` permutes the data and calculates the difference in permuted densities.  `B` is the number of bootstrap replicates to perform, `npoints` is the number of grid points in each direction.  It returns a 3D array.

```{r permdiff}
perm_dens <- permuteDensities(B = 15, npoints = n)
dim(perm_dens) # array of n*n*B dimensions

# plot of one permuted density
persp3D(x = 1:n, y = 1:n, z = perm_dens[, , 1])
```

This chunk of code then calculates the p-values at each grid point.

```{r pvals}
# calculating p-values from permutations
pvals <- matrix(NA, nrow = n, ncol = n) %>% 
  data.frame() %>% tbl_df()
for(i in 1:n) {
  for(j in 1:n) {
    pvals[i,j] <- pdata(abs(obs_diff[i, j]), 
                        abs(perm_dens[i, j, ]), 
                        lower.tail = F)
  }
}

# turning into a long data.frame
pvals_melt <- pvals %>% 
  gather(location, pval) %>% 
  mutate(col = rep(1:n, each = n), 
         row = rep(1:n, n)) %>% 
  select(row, col, pval)
```

Here we plot the observed p-values by location

```{r observed_pvals}
qplot(row, col, data = pvals_melt, colour = pval) + 
  scale_colour_gradient2(low = muted("red"), mid = "grey", high = muted("blue"), 
                         midpoint = .2)
```

And here we plot the p-values by whether they are greater than 0.05

```{r cutoff_pvals}
pvals_melt <- pvals_melt %>% 
  mutate(signif = ifelse(pvals < .05, "mouse", "elephant") %>% 
           factor(levels = c("mouse", "elephant")))
qplot(row, col, data = pvals_melt, colour = signif) + 
  labs(x = "Horizontal Location", y = "Vertical Location") + 
  
```


